---
title:  "The eR-Biostat initiative : Using R markdown to develop E-learning capacity"
author:
- name: Olajumoke Evangelina Owokotomo
  affiliation: Hasselt University, Belgium
subtitle:  Logistic regression modeling using R at the SUSAN-SSACAB 2019 Conference,  Cape Town, South Africa
date: September 8 - 11, 2019
output:
  rmdformats::readthedown:
    self_contained: no
    highlight: kate
    use_bookdown: TRUE
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
categories: ["R"]
bibliography: bibliography.bib
---



```{r style, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
suppressPackageStartupMessages({
  library(BiocStyle)
  library(knitr)
  library(rmarkdown)
  library(bookdown)
})
options(width = 100)
opts_chunk$set(fig.width = 5, fig.height = 5)

```

# The GLM Course {-}
## An example from the GLM course {-}
The Aspirin Use and Myocardial Infarction dataset includes the results of a study on the relationship between aspirin use and heart attacks by the Physicians’ Health Study
Research Group at Harvard Medical School [@ag]. The Physicians’ Health Study was a 5-year randomized study of whether regular aspirin intake reduces
mortality from cardiovascular disease. Every other day, physicians participating
in the study took either one aspirin tablet or a placebo. The study was
blinded-those in the study did not know whether they were taking aspirin or a
placebo. Of the 11034 physicians taking placebo, 189 suffered heart attacks, a proportion of $189/11034=0.0171$. Of the 11037 taking aspirin, 104 had heart attacks, a proportion of
$0.0094$. The R code below is used to construct the data. 

```{r,, eval = FALSE}
resp<-as.factor(c(rep(0,189),rep(1,10845),rep(0,104),rep(1 ,10933)))
trt<-as.factor(c(rep(1,189),rep(1,10845),rep(2,104),rep(2,10933)))
cbind(resp,trt)
Aspirin.1<-table(trt,resp)
Aspirin.1
row.names(Aspirin.1)=c("Placebo","Aspirin")
```

A barplot of the data can be produced using the `barplot()` function

```{r, eval = FALSE}
barplot(c(c(189,10845)/(189+10845),c(104,10933)/(104+10933))
,col=c(1,2,1,2)
,names=c("Yes","No","Yes","No"))
```

```{r, eval = TRUE, echo=FALSE}
barplot(c(c(189,10845)/(189+10845),c(104,10933)/(104+10933))
,col=c(1,2,1,2)
,names=c("Yes","No","Yes","No"))
```

# Model formulation {-}

Let $Y_{i}$ be the response variable, 
\[
Y_{i}=\left \{
\begin{array}{ll}
1, & \mbox{Myocardial Infarction}, \\
0, & \mbox{Healty}. 
\end{array}
\right .
\]

Our main interest it to model the probability
\[
P(Y_{i}=1)=\pi_{i},
\]

with linear predictor given by $\beta_{0}+\beta_{1}$Treatment$_{i}$.

The parameter $\beta_{1}$ is the log(odds ratio), that is $OR=\exp(\beta_{1})$.


The model can be estimated in R using the `glm()` function for the analysis we use the zero/one version of the data which creates two vectors of the data, that is

```{r, eval = FALSE}
resp<-as.factor(c(rep(1,189),rep(0,10845),rep(1,104),rep(0 ,10933)))
trt<-as.factor(c(rep(1,189),rep(1,10845),rep(2,104),rep(2,10933)))
```

```{r, eval = TRUE, echo=FALSE}
resp<-as.factor(c(rep(1,189),rep(0,10845),rep(1,104),rep(0 ,10933)))
trt<-as.factor(c(rep(1,189),rep(1,10845),rep(2,104),rep(2,10933)))
table(trt,resp)
```

Using the two vectors, `resp` and `trt` the model is fitted using the following code:  

```{r, eval = FALSE}
fit.myoc<-glm(resp~relevel(trt,2),family=binomial(link="logit"))
```

We model the probability of success `P(Z=1)`. This is the probability to have cardiovascular disease. We want to see if Aspirin intake has an effect on the probability to have Myocardial infarction.

The output is given below.
```{r, eval = TRUE, echo=FALSE}
resp<-as.factor(c(rep(0,189),rep(1,10845),rep(0,104),rep(1 ,10933)))
trt<-as.factor(c(rep(1,189),rep(1,10845),rep(2,104),rep(2,10933)))
fit<-glm(resp~relevel(trt,2),family=binomial(link="logit"))
summary(fit)
```

The parameter estimate for $\beta_{1}$ is equal to `0.605`


```{r, eval = TRUE, echo=FALSE}
fit$coeff
```

and the odds ratio is equal to 1.83, indication that taking one Aspirin tablet per day reduce the risk for Myocardial Infarction.

```{r, eval = TRUE, echo=FALSE}
OR<-exp(fit$coeff[2])
OR
```

# Session information {-}

```{r}
sessionInfo()
```

# References
